---
title: "Volcano Plot"
author: "Pumla"
date: "2024-06-18"
output: html_document
---

#Setting up environment 
#Clean environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = F, dplyr.summarise.inform = F) # avoid truncated output in R console and scientific notation
```


# Load libraries
```{r}
library(edgeR)
library(pheatmap)
library(dplyr)
library(ggplot2)
library(ggrepel)
```



# Set the working directory
```{r}
setwd("~/inktset1/Gene names retrived")
```


# Load count data
```{r}
count_data <- read.csv('Counts.csv', header = TRUE, row.names = 1)
```



# Load sample information
```{r}
sample_info <- read.csv('Design.csv', header = TRUE, row.names = 1)
```



# Set up the DGEList object
```{r}
dge <- DGEList(counts = count_data, group = sample_info$Treatment)
```



# Filter lowly expressed genes
```{r}
keep <- rowSums(cpm(dge) > 1) >= 1
dge <- dge[keep,]
```



# Estimate dispersion
```{r}
dge <- estimateGLMCommonDisp(dge)
dge <- estimateGLMTrendedDisp(dge)
dge <- estimateGLMTagwiseDisp(dge)
```


# Perform differential expression analysis
```{r}
fit <- glmQLFit(dge)
qlf <- glmQLFTest(fit, coef = 2)
```


# Extract results
```{r}
edger_result <- topTags(qlf, n = nrow(dge))
```

# Save the edgeR result
```{r}
write.table(edger_result$table, "edger_results.csv", sep = ",", col.names = NA)
```

# Read in the CSV file
```{r}
edger_results <- read.csv("edger_results.csv")
```

# Rename the first column from "X" to "ensembl_gene_id"
```{r}
colnames(edger_results)[1] <- "ensembl_gene_id"
```


# Check the column names to confirm the change
```{r}
colnames(edger_results)
```


# Read the CSV file into a data frame
```{r}

fpkm_data <- read.csv("FPKM.csv")
```


#checking to make sure both file have Gene id as first column
```{r}
colnames(edger_results)
```
colnames(edger_results)

# Function to remove versioned part from Ensembl gene IDs
```{r}
remove_version <- function(gene_ids) {
  pattern <- "^(\\w+)\\.\\d+$"  # Regular expression to match the part before the dot
  unversioned_ids <- sub(pattern, "\\1", gene_ids)  # Extract the part before the dot
  return(unversioned_ids)  # Return the unversioned gene IDs
}
```


# Apply the function to the ensembl_gene_id column in edger_results
```{r}
edger_results$ensembl_gene_id <- remove_version(edger_results$ensembl_gene_id)
```



#checking to make sure both file have Gene id as first column
```{r}
colnames(edger_results)
```



#now i want to assign gene names and gene type to my data set
#Load biomaRt library
```{r}
library(biomaRt)
```

#Use Ensembl biomart for mouse
```{r}
ensembl <- useEnsembl(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")  # for mouse
```


#Specify attributes to retrieve
```{r}
attributes <- c("ensembl_gene_id", "external_gene_name", "description")
```


#Retrieve gene information for mouse
```{r}
gene_info_mouse <- getBM(attributes = attributes, mart = ensembl)
```


#Write the gene_mapping_table to a CSV file
```{r}
write.csv(gene_info_mouse, file = "gene_info_mouse.csv", row.names = FALSE)
```


```{r}
colnames(edger_results)

colnames(gene_info_mouse)
```


#Assuming edger_results and gene_info_mouse dataframes are available

#Merge based on ensembl_gene_id
```{r}
merged_data <- merge(edger_results, gene_info_mouse, by = "ensembl_gene_id")
```


# Write the merged data to a CSV file
```{r}
write.csv(merged_data, file = "merged_data.csv", row.names = FALSE)
```


```{r}
colnames(merged_data)
```


# Sort merged_data by absolute log fold change (descending order)
```{r}
merged_data <- merged_data %>% 
  arrange(desc(abs(logFC)))
```


# Select the top 10 most differentially expressed genes
```{r}
top_10_genes <- head(merged_data, 10)
```


# Create Volcano Plot
```{r}
plot(ggplot(merged_data, aes(x = logFC, y = -log10(FDR))) +
  geom_point(color = ifelse(merged_data$FDR < 0.05, "red", "black"), alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  labs(x = "Log2 Fold Change", y = "-log10(FDR)", title = "Volcano Plot") +
  theme_minimal() +
  geom_text(data = filter(merged_data, FDR < 0.05), aes(label = external_gene_name), vjust = -0.5, color = "red"))
```


#Filter significant genes
```{r}
significant_genes <- merged_data %>% filter(FDR < 0.05)
```


#Select top 10 significant genes
```{r}
top_10_significant_genes <- head(significant_genes, 10)
```


#Create Volcano Plot
```{r}
plot(ggplot(merged_data, aes(x = logFC, y = -log10(FDR))) +
  geom_point(color = ifelse(merged_data$FDR < 0.05, "red", "black"), alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_text(data = top_10_significant_genes, aes(label = external_gene_name), vjust = -0.5, color = "red") +
  labs(x = "Log2 Fold Change", y = "-log10(FDR)", title = "Volcano Plot") +
  theme_minimal())
```


#Select top 5 significant genes
```{r}
top_5_significant_genes <- head(significant_genes, 5)
```


#Create Volcano Plot for Top 5 Significant Genes
```{r}
plot(ggplot(merged_data, aes(x = logFC, y = -log10(FDR))) +
  geom_point(color = ifelse(merged_data$FDR < 0.05, "red", "black"), alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_text(data = top_5_significant_genes, aes(label = external_gene_name), vjust = -0.5, color = "red") +
  labs(x = "Log2 Fold Change", y = "-log10(FDR)", title = "Volcano Plot for Top 5 Significant Genes") +
  theme_minimal())
```

